---
title: "Introduction to Air Quality Data Analysis"
#output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

# Overview

This .rmd file uses a cleaned .rdata file as its starting point - in this case 'roxbury_data_cleaned.rdata', which has the entire dataframe of all sensors in Roxbury. Generally, the code does this:

1. Loads packages
2. Loads the .rdata file
3. Filters by sensor or groups of sensors
4. Sanity checking via data tables and time series
5. Additional cleaning, as necessary
6. Plotting functions and exploratory data analysis

   

# Loading initial packages

```{r, warning= FALSE, error = FALSE}
# Check and install required packages if necessary
packages <- c("openair", "openairmaps", "leaflet", "dplyr", "chron", "timeDate", "data.table")
install.packages(packages[!sapply(packages, requireNamespace, quietly = TRUE)])

# Load required packages for data manipulation and analysis
invisible(sapply(packages, library, character.only = TRUE))

# Set options
knitr::opts_chunk$set(echo = FALSE, message = FALSE)

```

# AIR QUALITY DATA LOADING 
## Load Air Quality Data file.

```{r}
# Load data frame
load("./data/roxbury_df_filtered.RData")


```

## Filter by the sensor(s) you want to analyze
```{r}
# Choose a single sensor - change the SN to choose your sensor. Your individual dataframe will be called 'my_df,' which you can call later for plotting functions
my_df <- mod_met_filtered %>% 
      filter(mod_met_filtered$sn %in% c('MOD-PM-00222'))
#my_df <- mod_met_filtered %>% 
    #  filter(mod_met_filtered$sn,'MOD-00028')

###Below are different permutations of sensor groupings that you can use to analyze data together. For example, the sensors around Nubian Station are: MOD-00024, MOD-00028, MOD-PM-00222, and MOD-00025. Instead of filtering by a single SN, you'll filter by a vector of SNs (c('MOD-00024', 'MOD-00028', 'MOD-PM-00222', 'MOD-00025')).

# Nubian Station (MOD-00024, MOD-00028, MOD-PM-00222, MOD-00025)
# nubian_data  <- mod_met_filtered %>% 
#       filter(mod_met_filtered$sn == c('MOD-00024', 'MOD-00028', 'MOD-PM-00222', 'MOD-00025'))

# Commuter rail (MOD-PM-00221, MOD-PM-00224)
# train_data <- mod_met_filtered %>% 
#       filter(mod_met_filtered$sn == c('MOD-PM-00221', 'MOD-PM-00224'))

# School buses (MOD-PM-00212, MOD-PM-00214, MOD-PM-00223)
# school_bus_data <- mod_met_filtered %>% 
#       filter(mod_met_filtered$sn == c('MOD-PM-00214', 'MOD-PM-00223'))



# Street buses (MOD-PM-00231, MOD-PM-00213)
# street_bus_data <- mod_met_filtered %>% 
#       filter(mod_met_filtered$sn == c('MOD-PM-00231', 'MOD-PM-00213'))

# Malcolm X Park data
# malcolm_x_data <- mod_met_filtered %>% 
#       filter(mod_met_filtered$sn == 'MOD-PM-00217')


# Known construction sites
# construction_data <- mod_met_filtered %>% 
#       filter(mod_met_filtered$sn == c('MOD-00024', 'MOD-PM-00231'))


```

# CLEANING STEPS - already done for 'roxbury_df_filtered.rdata', but it's here if you need to do more cleaning.
```{r}
#outlierReplace Function
outlierReplace = function(dataframe, cols, rows, newValue = NA) {
    if (any(rows)) {
        set(dataframe, rows, cols, newValue)
    }
}

```


# Define threshold values
```{r}
pm10_threshold <- 1000
pm25_threshold <- 100
pm1_threshold <- 50
no_threshold <-150
no2_threshold <- 400
co_threshold <- 5000
o3_threshold <- 200
ws_threshold <- 10

```

# Replace values above threshold defined above, filter values below zero
```{r}

outlierReplace(my_df, "pm10", which(my_df$pm10 > pm10_threshold), NA)
outlierReplace(my_df, "pm25", which(my_df$pm25 > pm25_threshold), NA)
outlierReplace(my_df, "pm1", which(my_df$pm1 > pm1_threshold), NA)

outlierReplace(my_df, "no", which(my_df$no > no_threshold), NA)
outlierReplace(my_df, "no2", which(my_df$no2 > no2_threshold), NA)
outlierReplace(my_df, "co", which(my_df$co > co_threshold), NA)
outlierReplace(my_df, "o3", which(my_df$o3 > o3_threshold), NA)
outlierReplace(my_df, "ws", which(my_df$ws > ws_threshold), NA)

outlierReplace(my_df, "pm10", which(my_df$pm10 < 0), NA)
outlierReplace(my_df, "pm25", which(my_df$pm25 < 0), NA)
outlierReplace(my_df, "pm1", which(my_df$pm1 < 0), NA)

outlierReplace(my_df, "no", which(my_df$no < 0), NA)
outlierReplace(my_df, "no2", which(my_df$no2 < 0), NA)
outlierReplace(my_df, "co", which(my_df$co < 0), NA)
outlierReplace(my_df, "o3", which(my_df$o3 < 0), NA)

```



# Summary statistics

```{r}
# Looking briefly at summary statistics will give you a sense of the variables that you just loaded. This can also tip you off to things that might need attention. 
# What do you notice in these summary statistics?

summary(my_df)

```

# Date formatting

```{r}
# Dates and times have a bunch of different possible formats. These formats can be the bane of your analysis existence. Thankfully the lubridate package (which we loaded above) makes things smoother. It even automatically detects daylight saving time!
my_df$date <- as.POSIXct(strptime(my_df$timestamp_local.x, format = "%Y-%m-%d %H:%M:%S", tz = "America/New_York"))


```

# Time series

```{r}
# We'll spend a lot of time working with the openair package in our project and will get a more in-depth introduction in the next class. But we'll draw on one of its functions now.
# As you look at this time series, what sticks out to you?
# Where do you notice points that seem outside of what is reasonable?
# Do you need to do additional cleaning, or do things look good?


timePlot(my_df, pollutant = "pm1")
timePlot(my_df, pollutant = "pm25")
timePlot(my_df, pollutant = "pm10")
#timePlot(aq_df, pollutant = c("co", "o3", "no", "no2", "pm1", "pm25", "pm10"), y.relation= "free")
#timePlot(aq_df, pollutant = c("co", "o3", "no", "no2"), y.relation= "free"). #error? does your sensor have gas-phase data or PM only? 
timePlot(my_df, pollutant = c("pm1", "pm25", "pm10"), main = "Amount of Consentration of PM1,PM25,PM10 from July 2022 to July 2023", y.relation= "free")



```


# EXPLORATORY DATA ANALYSIS

## Temporal variability: Diurnal and annual/seasonal profiles

```{r}
# Diurnals: First look at a "typical day" for different pollutants
# Read openair manual to find different ways to use this function. Things that may be helpful: selectByDate; type (season, etc); normalise; subset


# Basic diurnal plots:
#timeVariation(my_df, pollutant = c("no", "no2", "co"), local.tz= "America/New_York", normalise = TRUE) #error? does your sensor have gas-phase data, or PM only?
timeVariation(my_df, pollutant = c("pm1", "pm25", "pm10"), normalise = TRUE, local.tz= "America/New_York") 

# If you want to select by certain months:
 timeVariation(selectByDate(my_df, month = c(1:5, 9:12)), pollutant = c('pm1', 'pm25', 'pm10'))

# If you want to create just a subset of the timeVariation output (diurnals only; seasonal only, etc)
 myOutput <- timeVariation(my_df, pollutant = "pm1", local.tz= "America/New_York", col = "firebrick", type = "season")#, subset = "hour") 
 plot(myOutput, subset = "hour")

```

## Directional analysis of pollutants

### Create polar plots (and other things in that family)

```{r}
# this group of functions allows you to explore the relationship between wind speed, wind direction, and pollutant concentrations. They all give slightly different flavors of the relationship between pollutants and meteorology, so explore and see what stories emerge. 

# There are a bunch of different things you can define here - type (seasonal, etc), statistic (median, mean, max, standard deviation, etc), color scale limits, icon size, figure transparency, etc. In particular, the type and statistic inputs give you *very* different pictures of what's happening in an environment.

polarPlot(my_df, pollutant = "pm1",main="Ploar Plot PM1")
polarPlot(my_df, pollutant = "pm1", type = "season", main="Ploar Plot PM1 by Season")
polarPlot(my_df, pollutant = "pm1", limits = c(0,6), type = "season",main="Ploar Plot PM1 by Season with Limit")
polarFreq(my_df, pollutant = "pm1",main="Polar Freq PM1")
pollutionRose(my_df, pollutant = "pm1",main="Polar Rose PM1")
polarAnnulus(my_df, pollutant = "pm1",main="Polar Anual PM1")
polarCluster(my_df, pollutant = "pm1", n.clusters = 4,main="Polar Cluster PM1")

polarPlot(my_df, pollutant = "pm25",main="Polar Plot PM25")
polarPlot(my_df, pollutant = "pm25", type = "season",main="Polar Plot PM25 by Season")
polarPlot(my_df, pollutant = "pm25", limits = c(0,6), type = "season",main="Polar Plor PM25 with Limit")
polarFreq(my_df, pollutant = "pm25",main="Polar Freq PM25")
pollutionRose(my_df, pollutant = "pm25",main="Polar Rose PM25")
polarAnnulus(my_df, pollutant = "pm25",main="Polar Anual PM25")
polarCluster(my_df, pollutant = "pm25", n.clusters = 4,main="Polar Cluster PM25")

polarPlot(my_df, pollutant = "pm10",main="Polar Plot PM10")
polarPlot(my_df, pollutant = "pm10", type = "season",main="Polar Plot PM10 by Season")
polarPlot(my_df, pollutant = "pm10", limits = c(0,21), type = "season",main="Polar Plot PM10 by Season with Limit")
polarFreq(my_df, pollutant = "pm10",main="Polar Freq PM10")
pollutionRose(my_df, pollutant = "pm10",main="Polar Rose PM10")
polarAnnulus(my_df, pollutant = "pm10",main="Polar Anual PM10")
polarCluster(my_df, pollutant = "pm10", n.clusters = 4,main="Polar Cluster PM10")

```

### Create Polar map plots

```{r}
#polarMap is an extension of polarPlot that places a polarPlot onto a map as an overlay. polarPlot is in a family of plots that include pollutionRose, polarAnnulus, polarFreq, percentileRose, and others that give slightly different visualizations of the same combined pollutant/met data. You should play with some of these too.

# There are a bunch of different things you can define here - type (seasonal, etc), statistic (median, mean, max, standard deviation, etc), color scale limits, icon size, figure transparency, etc. In particular, the type and statistic inputs give you *very* different pictures of what's happening in an environment.

polarMap(my_df,
         pollutant = "pm1", 
         key.position = "bottom",
         key.header = "PM1 (ug/m3)", 
         key.footer = NULL, 
         x = "ws",
         latitude = "lat",
         longitude = "lon", 
         provider = "OpenStreetMap",
         cols = "jet",
         alpha = 0.8,
         key = TRUE,
         iconWidth = 200,
         iconHeight = 200,
         fig.width = 4,
         fig.height = 4
        )


# polarMap(my_df,
#          pollutant = "no", 
#          key.position = "bottom",
##          key.header = "NO (ppb)", 
#        key.footer = NULL, 
 #         x = "ws",
  #        latitude = "lat",
   #       longitude = "lon", 
    #      provider = "OpenStreetMap",
     #     limits = c(0, 4),
      #    cols = "jet",
      #    alpha = 0.8,
      #    key = TRUE,
      #    iconWidth = 200,
      #    iconHeight = 200,
      #    fig.width = 4,
      #    fig.height = 4
      #    )
# 
# 
# polarMap(my_df,
#          pollutant = "no", 
#          key.position = "bottom",
#          key.header = "NO (ppb)", 
#          x = "ws",
#          latitude = "lat",
#          longitude = "lon", 
#          provider = "OpenStreetMap",
#          limits = c(0, 4),
#          cols = "jet",
#          key = TRUE,
#          )

```

### Create some calendar plots to explore long-term tempral trends

```{r}
# These can help identify specific dates or ranges of dates that had particularly high pollutant concentrations. You may want to look more closely at conspicuously elevated pollutant days.

calendarPlot(my_df, pollutant = "pm1")
calendarPlot(my_df, pollutant = "pm1", limits = c(0,80))
# 
# 
# calendarPlot(my_df, pollutant = "pm10", annotate = "value", limits = c(0,80),
# lim =50, cols = "Purples", col.lim = c("black", "orange"), layout = c(4, 3))
# 
# trendLevel(my_df, pollutant = "pm10", limits = c(0,20))


```
